Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
Incorporate feedback on Imputation Strategies section

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation (\cite{witten2016data}).

The algorithm used by the \textit{Impute} operator proceeds by iteratively fitting regression trees to a subset of the
data and the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
improves as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
across epochs) or a fixed number of epochs is reached.

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset
(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.

Given the nature of the imputation strategy used here, {\it Impute} operators are blocking
in our system. An extension of \ProjectName{} could consider regression algorithms that learn in
an online nature; this would reduce the cost of the imputation at the possible expense of
quality.

>>>>

\ProjectName{} is designed so that any imputation strategy, typically a predictive machine
learning model,
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
An imputation strategy is a function $I(T, C) \rightarrow T'$ which fills in the missing
attributes $C$ in a set of tuples $T$, returning a new set of tuples $T'$.
In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation~\cite{witten2016data}.

The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
regression trees to a subset of the data and the target imputation attribute. In each
iteration, the missing values of a single attribute are replaced with newly imputed values.
In one {\it epoch} of the algorithm, each of the target imputation
attributes is filled in with imputed values in a single iteration.
With each {\it epoch}, the quality of the imputation improves as values progressively
reflect more accurate relationships between attributes. The algorithm terminates when
convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
number of epochs is reached.

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be another avenue of assessing
imputation confidence.
